---
title: "Homework Assignment 1"
author: "Alexander Reiner"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
  pdf_document:
    toc: true
---

```{r}
# Data Manipulation
nyc <- read.csv("nycflights.csv")

# Delete the column named "X"
nyc$X <- NULL

#Recode columns as factors as needed
nyc$carrier <- as.factor(nyc$carrier)
nyc$flight <- as.factor(nyc$flight)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)

#Add airgain column
nyc$airgain <- c(nyc$dep_delay - nyc$arr_delay)
```


# Data Exploration

### How Many Airlines are there?
```{r, results= FALSE}
# use length and levels function
length(levels(nyc$carrier))
```
There are `r length(levels(nyc$carrier))` airlines

### How many flights there were by the airline code oo?
```{r, results= FALSE}
# Use nrow
nrow(nyc[nyc$carrier == "OO",])
```
There are `r nrow(nyc[nyc$carrier == "OO",])` flights from airline code OO.

### How long is the shortest flight out of any NYC airport?
```{r, results= FALSE}
# Use min function

min(nyc$air_time, na.rm = TRUE)
```
The shortest flight out of any NYC airport is `r min(nyc$air_time, na.rm = TRUE)` minutes.

### How many flights were there by United Airlies on January 12th 2013?
```{r}
#Use nrow

nrow(nyc[nyc$carrier == "UA" & nyc$year == 2013 & nyc$month == 1 & nyc$day ==12, ])
  
```
On January 12, 2013, there were `r nrow(nyc[nyc$carrier == "UA" & nyc$year == 2013 & nyc$month == 1 & nyc$day ==12, ])` flights by United Airlines.

# Arrival Delay

### What was the average arrival delay for all airports and all airlines combined in January 2013?
```{r}
# Use mean function
jan2013 <- nyc[nyc$month == 1, ]

mean(jan2013$arr_delay, na.rm=TRUE)

```
The average arrival delay for all airports and airlines in January 2013 is `r mean(jan2013$arr_delay, na.rm=TRUE)` minutes.

### What was the median arrival delay for all airports and all airlines combined in January 2013?
```{r, results= FALSE}
median(jan2013$arr_delay, na.rm = TRUE)

```
The median arrival delay for all airports and airlines in January 2013 is `r median(jan2013$arr_delay, na.rm = TRUE)` minutes.

# Airline Performance

### Average Arrival Delay by Airline

```{r}
# Use function tapply

arr_delay_by_airline <- round(tapply(nyc$arr_delay, nyc$carrier, mean,  na.rm=TRUE), 2)

arr_delay_by_airline

```

### Arrival Delay Barplot

```{r}
library(RColorBrewer)
color <- brewer.pal(16, "Set1")

barplot(sort(arr_delay_by_airline, decreasing = TRUE), xlab="Airline", col.lab = "red", ylab="Minutes", col=color, main= "Average Arrival Delays by Airline", cex.names = 0.53)
```

### Airline with the Highest Average Arrival Delay

`r names(arr_delay_by_airline[max(arr_delay_by_airline) == arr_delay_by_airline])` airline has the highest average arrival delay.

### Airline with the Lowest Average Arrival Delay

`r names(arr_delay_by_airline[min(arr_delay_by_airline) == arr_delay_by_airline])` airline has the lowest average arrival delay.

### Are there airlines that have negative average delay?

Yes, there are two airlines: `r names(arr_delay_by_airline[(arr_delay_by_airline) <= 0])`.

# Air Gain
```{r}
nyc$airgain <- c(nyc$dep_delay - nyc$arr_delay)
```

### Air Gain Statistics
```{r, results= FALSE}
round(mean(nyc$airgain, na.rm=TRUE), 2)
max(nyc$airgain, na.rm=TRUE)
min(nyc$airgain, na.rm=TRUE)
```

The average time gained is `r mean(nyc$airgain, na.rm=TRUE)` minutes.
The most time gained on a single flight is `r max(nyc$airgain, na.rm=TRUE)` minutes.
The least time gained on a single flight is `r min(nyc$airgain, na.rm=TRUE)` minutes.

### Air-Time Gained By Month
```{r}
# Use function tapply
airgain_by_distance <- round(tapply(nyc$airgain, nyc$month, mean,  na.rm=TRUE), 2)

airgain_by_distance

barplot(airgain_by_distance, xlab="Month", col.lab = "red", ylab="Minutes", col=color, main= "Average Time Made Up from Delays by Month", cex.names = 0.53)
```

### Air-Time Gained by Origin Airport
```{r}
# Use function tapply
airgain_by_origin <- round(tapply(nyc$airgain, nyc$origin, mean,  na.rm=TRUE), 2)

airgain_by_origin

barplot(airgain_by_origin, xlab="Origin Airport", col.lab = "red", ylab="Minutes", col=color, main= "Average Time Made Up by Origin Airport", cex.names = 0.53)
```

### Do airlines actually gain any time when in air on average?
Yes, airlines gain on average `r round(mean(nyc$airgain, na.rm = TRUE), 2)` minutes on average.

### Average Airgain for Airlines
```{r}
# Use function tapply
airgain_by_airline <- round(tapply(nyc$airgain, nyc$carrier, mean,  na.rm=TRUE), 2)

airgain_by_airline
```

### Which airline made up the most amount of time while in air?
`r names(airgain_by_airline[max(airgain_by_airline) == airgain_by_airline])` made up the most time while in air by `r max(airgain_by_airline, na.rm = TRUE)` minutes on average.

### Which airline made up the least amount of time while in air?
`r names(airgain_by_airline[min(airgain_by_airline) == airgain_by_airline])` made up the least time while in air by `r min(airgain_by_airline, na.rm = TRUE)` minutes on average.

### Airgain by Airline
```{r}
barplot(sort(airgain_by_airline, ), xlab="Airline", col.lab = "red", ylab="Minutes", col=color, main= "Average Time Made Up from Delays by Airline", cex.names = 0.53)
```

